---
title: "PLPPM 5020: Temporal Progress"
output: 
  html_document:
    number_sections: true 
    df_print: paged
    theme: cosmo
    toc: true
    toc_float:
      collapsed: yes
    self_contained: yes
---

# Libraries
First we need to run (and if necessary, install) some packages that we are going to use for these analyses

```{r message=FALSE, warning=FALSE}
# This code check if you have the package installed in R
if (!require(epifitter)) {
  install.packages("epifitter")
}

if (!require(ggthemes)) {
  install.packages("ggthemes")
}

if (!require(cowplot)) {
  install.packages("cowplot")
}

```


```{r message=FALSE, warning=FALSE}
# This code load the packages we'll be using
library(epifitter)
library(tidyverse)
library(ggthemes)

#This code set the graph theme we`ll be using throughout the script
theme_set(cowplot::theme_half_open())
```




# Simulating disease progress curves

The `sim_` functions, regardless of the model, require the same set of six arguments. By default, at least two arguments are required (the others have default values)

-   `r`: apparent infection rate
-   `n`: number of replicates

When `n` is greater than one, replicated epidemics (e.g. replicated treatments) are produced and a level of noise (experimental error) should be set in the `alpha` argument. These two arguments combined set will generate `random_y` values, which will vary randomly across the defined number of replicates.

The other arguments are:

-   `N`: epidemic duration in time units
-   `dt`: time (fixed) in units between two assessments
-   `y0`: initial inoculum
-   `alpha`: noise parameters for the replicates



## Models

### Exponential

Let's simulate a curve resembling the exponential growth.

#### Simulation



```{r}
exp_data = sim_exponential(N = 100,      # Total time course of the epidemic
                           dt = 1,      # Time intervals
                           y0 = 0.01,   # Initial inoculum
                           r = 0.1,     # Apparent infection rate
                           n = 1,       # Number or replicates or sample size for each time step
                           alpha = 0.2  # Variation parameter
                           )
```

A `data.frame` object is produced with four columns:

-   `replicates`: the curve with the respective ID number
-   `time`: the assessment time
-   `y`: the simulated proportion of disease intensity
-   `random_y`: randomly simulated proportion disease intensity based on the noise

```{r}
exp_data
```


##### Graph

Use the [`ggplot2`](https://ggplot2.tidyverse.org/) package to build impressive graphics!

```{r}
exp_data %>% 
  ggplot(aes(time, y)) +
  geom_point()
```






### Monomolecular

The logic is exactly the same here.

#### Simulation
```{r}
mono_data = sim_monomolecular(N = 100,      # Total time course of the epidemic
                              dt = 1,      # Time intervals
                              y0 = 0.01,   # Initial inoculum
                              r = 0.1,     # Apparent infection rate
                              K = 1,       # Maximum asymptote 
                              n = 1,       # Number or replicates or sample size for each time step
                              alpha = 0.2  # Variation parameter
                              )
mono_data

```

#### Graph

```{r}
mono_data %>% 
  ggplot(aes(time, y)) +
  geom_point()
```



### Logistic
#### Simulation
```{r}
logi_data = sim_logistic(N = 100,      # Total time course of the epidemic
                         dt = 1,      # Time intervals
                         y0 = 0.01,   # Initial inoculum
                         r = 0.1,     # Apparent infection rate
                         K = 1,       # Maximum asymptote 
                         n = 1,       # Number or replicates or sample size for each time step
                         alpha = 0.2  # Variation parameter
                         )
logi_data
```


#### Graph
```{r}
logi_data %>% 
  ggplot(aes(time, y)) +
  geom_point()
```





### Gompertz
#### Simulation
```{r}
gompi_data = sim_gompertz(N = 100,      # Total time course of the epidemic
                          dt = 1,      # Time intervals
                          y0 = 0.01,   # Initial inoculum
                          r = 0.1,     # Apparent infection rate
                          K = 1,       # Maximum asymptote 
                          n = 1,       # Number or replicates or sample size for each time step
                          alpha = 0.2  # Variation parameter
                         )
gompi_data
```

#### Graph
```{r}
gompi_data %>% 
  ggplot(aes(time, y)) +
  geom_point()
```



### Comparison

```{r}
data_models = bind_rows(Esponential = exp_data,
                        monomolecular = mono_data,
                        Logistic = logi_data,
                        Gompertz = gompi_data,
                        .id = "model")
```


```{r}
data_models %>% 
  ggplot(aes(time, y, color = model))+
  geom_point()+
  ylim(0,1)
```



## Interact

[Click here](https://alvesks.shinyapps.io/simDPC/){target="_blank"}




# Fitting models

## Data
```{r}
data("PowderyMildew")
PowderyMildew
```

### Visualization

```{r}
PowderyMildew %>% 
  ggplot(aes(time,sev, color = irrigation_type))+
  geom_point()+
  facet_wrap(~moisture)
```



## One disease progress curve

### Filter just one treatment
```{r}
single_curve = PowderyMildew %>%
  filter(irrigation_type == "Overhead",
         moisture == "Moderate moisture") 
single_curve
```

#### Visualization
```{r}
single_curve %>% 
  ggplot(aes(time, sev))+
  geom_point()
```

### Fitting

### Linear regression
```{r}
linear_fit = fit_lin(time = single_curve$time,
                     y = single_curve$sev)
linear_fit
```


```{r fig.height=4, fig.width=8}
plot_fit(linear_fit)
```

### Non-linear regression

```{r}
nonlinear_fit = fit_nlin(time = single_curve$time,
                         y = single_curve$sev)
nonlinear_fit
```

```{r fig.height=4, fig.width=8}
plot_fit(nonlinear_fit)
```


### Estimate maximun assimptote

```{r}
assimp_fit = fit_nlin2(time = single_curve$time,
                       y = single_curve$sev,
                       starting_par = list(y0 = 0.01, r = 0.03, K = 0.2))
assimp_fit
```

```{r fig.height=4, fig.width=8}
plot_fit(assimp_fit)
```



## Multiple DPCs

```{r}
fit_pd = fit_multi(time_col = "time",
          intensity_col = "sev",
          strata_cols = c("irrigation_type", "moisture"),
          estimate_K = T,
          nlin = T,
          data = PowderyMildew)

```

### Model parameters
```{r}
fit_pd$Parameters
```

### Model data

```{r}
fit_pd$Data
```

### Visualization

```{r fig.height=4, fig.width=8}
fit_pd$Data %>% 
  filter(model == "Gompertz") %>% 
  ggplot()+
  geom_point(aes(time, y, color= irrigation_type))+
  geom_line(aes(time, predicted, color= irrigation_type))+
  facet_wrap(~moisture)
```


# Area under disease progress curves

## Single DPC
```{r}
AUDPC(time = single_curve$time,
      y = single_curve$sev)
```

## Multiple DPCs
```{r}
PowderyMildew %>% 
  group_by(irrigation_type, moisture,block ) %>% 
  summarise(audpc = AUDPC(time = time, y = sev),
            audps = AUDPS(time = time, y = sev))
```




```{r fig.height=4, fig.width=8}
PowderyMildew %>% 
  group_by(irrigation_type, moisture,block ) %>% 
  summarise(audpc = AUDPC(time = time, y = sev),
            audps = AUDPS(time = time, y = sev)) %>% 
  ggplot(aes(irrigation_type, audpc, color = moisture)) + 
  geom_boxplot()
```

```{r fig.height=4, fig.width=8}
PowderyMildew %>% 
  group_by(irrigation_type, moisture,block ) %>% 
  summarise(audpc = AUDPC(time = time, y = sev),
            audps = AUDPS(time = time, y = sev)) %>% 
  ggplot(aes(irrigation_type, audps, color = moisture)) + 
  geom_boxplot()
```


